Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-05 latex/main.tex
Reshuffled some text

The ``succinct tree sequence'' (STS) is an ARG data structure
that is both general (in terms of the types of ancestry that can be described)
and computationally efficient \cite{Wong2024}.
Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016},
STS has been extended and applied to
forward-time simulations \cite{Kelleher2018,Haller2019},
calculation of population genetics statistics \cite{Ralph2020} and
ARG inference \cite{Kelleher2019,Wohns2022}.
The STS is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations.
A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor.
The genetic inheritance between a pair of nodes along a segment of genome
is defined by the edge ($l$, $r$, $p$, $c$),
which states that child node $c$ inherited its genome from parent node $p$
from left coordinate $l$ to right coordinate $r$.
A site defines a position on the genome and the ancestral state (allele) at that site.
A mutation records the site and node IDs where a mutation occurs and the derived state (allele).

Given the node, edge, site and mutation tables,
we can efficiently construct the local genealogical trees along the genome (arising from recombination)
and perform a range of calculations efficiently
by reasoning about the differences between these local trees \cite{Kelleher2016,Ralph2020}.
These algorithms have led to performance increases of several orders of magnitude over
the state-of-the-art in a range of applications \cite{Kelleher2016,Kelleher2018,Kelleher2019,Ralph2020,Baumdicker2022}.
The STS encoding is also very concise, allowing, for example,
for millions of complete human genomes to be stored in a few GB of space \cite{Kelleher2019},
and so it is particularly well suited to representing large SARS-CoV-2 genealogies.

The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed
using the mature and feature-rich tskit library \cite{Kelleher2018,Ralph2020}.
Tskit is widely used in population genetics, and has been incorporated into many downstream applications
(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2022,Mahmoudi2022,Petr2022,Rasmussen2022,Zhang2023}).

@@

During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates.
We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}.
We skipped low-quality sequences that had more than 500 missing sites
(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
We also skipped ~40,000 samples which had a collection date of December 31, 2020.
The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
We suspect that this batch of samples might be unusually enriched for wrong collection dates,
likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

@@

\subsubsection*{Identification of robust recombination events}
[TODO update refs!]

@@

We used Pango lineage designations to select \textit{representative} samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an "origination node" for a focal Pango lineage,
defined as the oldest internal node in the ARG
whose descendants in any tree include more than 50\% of the Pango samples of the focal type,
and where the internal node is itself designated as the focal Pango.
The representative sample is then chosen as the oldest sample under the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We simplify the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e. the Pango Xs) and three "singleton recombinant samples"
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

@@

The simplified ARG contains four single-breakpoint recombination nodes,
corresponding to the origins of Delta, BA.2, BA.5, and a further BA.2 recombinant
which we deem artifactual by our standard criteria
and treat as non-recombinant for the purposes of plotting
(the breakpoint is on the far right at position 27382,
and supported by only 4 distinguishing sites / 3 loci).
To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
and extract non-recombinant structures from the ARG
by creating a "base tree" of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
a Delta subtree of 160 Pangos,
a BA.2 subtree of 151 sample (excluding BA.5 descendants) and
a BA.5 subtree of 251 samples.
Figure SX shows each of these trees compared against the equivalent dated UShER topology,
with the left and right parents of Delta and BA.2 highlighted in the base tree.
The order of tips in the two trees was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
The entire pipeline is saved as a notebook at URL (?).

@@

\subsubsection*{Counting supporting loci for recombinant parents}
When counting such loci, we considered nearly adjacent sites (separated by at most three bases) as single loci,
to avoid over-counting support from correlated loci, which would lead to artificial signals of recombination.
Also, when counting supporting loci for a focal parent,
we penalized loci where the character(s) in the recombinant matches that in the non-focal parent but not in the focal parent.

>>>>

The ``succinct tree sequence'' (STS) is an ARG data structure
that is both general (in terms of the types of ancestry that can be described)
and computationally efficient \cite{Wong2024}.
Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016},
STS has been extended and applied to
forward-time simulations \cite{Kelleher2018efficient,Haller2019},
calculation of population genetics statistics \cite{Ralph2020} and
ARG inference \cite{Kelleher2019,Wohns2022}.
The STS is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations.
A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor.
The genetic inheritance between a pair of nodes along a segment of genome
is defined by the edge ($l$, $r$, $p$, $c$),
which states that child node $c$ inherited its genome from parent node $p$
from left coordinate $l$ to right coordinate $r$.
A site defines a position on the genome and the ancestral state (allele) at that site.
A mutation records the site and node IDs where a mutation occurs and the derived state (allele).

Given the node, edge, site and mutation tables,
we can efficiently construct the local genealogical trees along the genome (arising from recombination)
and perform a range of calculations efficiently
by reasoning about the differences between these local trees \cite{Kelleher2016,Ralph2020}.
These algorithms have led to performance increases of several orders of magnitude over
the state-of-the-art in a range of applications
\cite{Kelleher2016,Kelleher2018efficient,Kelleher2019,Ralph2020,Baumdicker2022}.
The STS encoding is also very concise, allowing, for example,
for millions of complete human genomes to be stored in a few GB of space \cite{Kelleher2019},
and so it is particularly well suited to representing large SARS-CoV-2 genealogies.

The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed
using the mature and feature-rich tskit library \cite{Kelleher2018efficient,Ralph2020}.
Tskit is widely used in population genetics, and has been incorporated into many downstream applications
(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2022,Mahmoudi2022,Petr2022,Rasmussen2022,Zhang2023}).

@@

During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates.
We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}.
We skipped low-quality sequences that had more than 500 missing sites
(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
We also skipped $\sim$40,000 samples which had a collection date of December 31, 2020.
The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
We suspect that this batch of samples might be unusually enriched for wrong collection dates,
likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

@@

\subsubsection*{Identification of robust recombination events}
[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.]

% This para was *{Counting supporting loci for recombinant parents}

When counting such loci, we considered nearly adjacent sites (separated by at
most three bases) as single loci, to avoid over-counting support from
correlated loci, which would lead to artificial signals of recombination. Also,
when counting supporting loci for a focal parent, we penalized loci where the
character(s) in the recombinant matches that in the non-focal parent but not in
the focal parent.

@@

We used Pango lineage designations to select representative samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an "origination node" for a focal Pango lineage,
defined as the oldest internal node in the ARG
whose descendants in any tree include more than 50\% of the Pango samples of the focal type,
and where the internal node is itself designated as the focal Pango.
The representative sample is then chosen as the oldest sample descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e. the Pango Xs) and three ``singleton recombinant
samples''
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

@@

The simplified ARG contains four single-breakpoint recombination nodes,
corresponding to the origins of Delta, BA.2, BA.5, and a further BA.2 recombinant
which we deem artifactual by our standard criteria
and treat as non-recombinant for the purposes of plotting
(the breakpoint is on the far right at position 27382,
and supported by only 4 distinguishing sites / 3 loci).
To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
and extract non-recombinant structures from the ARG
by creating a ``base tree'' of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
a Delta subtree of 160 Pangos,
a BA.2 subtree of 151 sample (excluding BA.5 descendants) and
a BA.5 subtree of 251 samples.
Figure SX shows each of these trees compared against the equivalent dated UShER topology,
with the left and right parents of Delta and BA.2 highlighted in the base tree.
The order of tips in the two trees was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
The entire pipeline is saved as a notebook at URL (?).

@@
